Haploid Selection, Sex Ratio Bias, and Transitions Between Sex-Determination Systems

### 2018-03-26 LaTeX/response.tex

\noindent\paragraph{Response}
\textcolor{blue}{We hope the responses below help...}

\noindent\subsubsection{}
The works of Kozielska 2014 and Ubeda 2015 model the combined effects of both haploid and diploid selection on the evolution of new sex-chromosomes. The work of [SOO] is thus not novel in considering both haploid and diploid selection on the evolution of new sex-chromosomes. It does generalize previous work by considering sex-specific viability selection and get some nice analytical results.

\noindent\paragraph{Response}
\textcolor{blue}{We now ...}

\noindent\subsubsection{}
The findings in [SOO] claimed to be surprising (``Surprisingly, we find that neither force (sex ratio selection nor associations with genes that have sex-specific effects) dominates the spread of new sex-determining systems alone.") ... [but this result is rather similar to those in Ubeda et al 2015].

\noindent\paragraph{Response}
\textcolor{red}{Here, we were referring to the result that $\lambda_{W'}^{XY} = \lambda_{Y'}^{ZW}$ with haploid selection, even when sex ratio biases evolve to become stronger or weaker (e.g., $\lambda_{Y'}^{ZW}$ vs. $\lambda_{W'}^{XY}$ with $r<R$ and male meiotic drive) or when sex ratio biases are present or absent (e.g., $\lambda_{W'}^{XY}$ vs. $\lambda_{Y'}^{ZW}$ with $r=1/2$ and male meiotic drive). The Ubeda et al. (2015) step 1 [?] results show that associations with a haploid selected locus can favour a new sex-determining allele despite causing biased sex ratios. Nevertheless, it was difficult for us to intuit whether the presence/absence of ancestral sex ratio bias would have a negligible or overwhelming effect on the evolution of new sex-determining alleles. For example, the second stage of Ubeda et al. (2015) involves the spread of a autosomal meiotic drive suppressor driven by fisherian sex ratio selection [CHECK SEX RATIO SELECTION MENTIONED] and Kozielska et al. show that ancestral sex ratio bias can favour new sex determining systems. We found our result, that these forces are equal, such that there is no difference between the propensity for XY-ZW or ZW-XY transitions with selection weak relative to recombination, to be surprising.} \textcolor{blue}{To clarify our meaning... }

\paragraph{} \noindent 
\noindent and unexpected (``Even more unexpectedly, we find that, to spread, new sex-determining alleles do not necessarily have to arise in closer linkage with genes that are differentially selected in males and females"), are rather similar to those in Ubeda et al 2015. This is a problem that repeats itself over the paper

\noindent\paragraph{Response}
\textcolor{red}{The Ubeda et al 2015 results do indeed show that differential selection between male and female diploids is not required for new sex-determining alleles to spread; they show that a new genetic sex-determining allele can spread from ancestral ESD when it is linked to a locus that experiences sex-specific meiotic drive. 
In this sentence, we were referring differential male/female selection in diploids \textit{or} the haploids produced by males/females. 
That is, we mean to say that our results demonstrate that new sex determining alleles can spread when they are \textit{less closely linked or completely unlinked} to loci under differential male/female haploid and/or diploid selection. 
To our knowledge, this has not been predicted by any verbal or mathematical models and we found this result to be very unexpected. }
\textcolor{blue}{We now ...}

@@

\noindent\paragraph{Response}
\textcolor{blue}{We now ...}

\noindent\subsubsection{}
Lines 68-69: ``If we assume that haploid selection at any particular locus predominantly occurs in one sex (e.g., meiotic drive during spermatogenesis), then such loci experience a form of sex-specific selection."

This is a stretch and creates confusion. I consider selection acting on individuals with different sex sex-specific selection, now when selection is acting on gametes (which are neither males or females) males or females are just different environments in which selection may take place. Neither the dynamics nor the constraints are the same.

\noindent\paragraph{Response}
\textcolor{red}{In this study, we are referring to organisms, such as animals and angiosperms, in which the type of gamete depends on the sex of the diploid that produced it. In this respect, one could consider gametic selection among sperm to be a form of male-specific selection. However, as noted here, the dynamics of selection during gametic selection cannot be captured by a diploid selection coefficient, which is why we model selection upon haploid genotypes specifically in two ways. [could tie this in with an expanded explanation of the mating system and the way we model haploid selection.]}
\textcolor{blue}{We now ...}

@@

\noindent\subsubsection{}
Lines 70-71: ``In this respect, we might expect that haploid selection would affect transitions between sex-determining systems in a similar manner to sex-specific diploid selection (as explored in [31,32])"

Following up my previous comment Ubeda 2015 explores this and shows that both forms of selection are different. For example sex-antagonistic selection is needed for the invasion of a new sex-determining gene with diploid selection while antagonistic drive in each of the sexes is not needed for the invasion of a new sex-determining gene with haploid selection.

\noindent\paragraph{Response}
\textcolor{red}{but \textit{differences} in haploid selection between the sexes are needed (i.e., drive in only one sex, maintained by ploidally-antagonistic selection). If $\alpha^{m}=\alpha^{f}$ (and all diploid male/female selection terms equal), then the new sex-determining allele does not invade. So I wouldn't use this argument to demonstrate that the two forms of selection are different. 
In this sentence, I was just trying to introduce the 'sex-specific' aspect of haploid selection in precisely the way Ubeda et al (2015) considers it. i.e., associations with alleles favoured in one sex (or the gametes produced by that sex) should favour a new sex-determining allele. Would have helped if I'd cited though... }

### 2018-03-26 LaTeX/main_revision.tex
first pass at all revisions done

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Haploid selection, sex ratio bias, and transitions between sex-determining systems} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Michael Francis Scott\textsuperscript{1\Yinyang*},
Matthew Miles Osmond\textsuperscript{2\Yinyang},
Sarah Perin Otto\textsuperscript{2}
\\
\bigskip
\textbf{1} UCL Genetics Institute, Department of Genetics, Evolution and Environment, University College London, Gower Street, London WC1E 6BT 
\\
\textbf{2} Department of Zoology, University of British Columbia, \#4200 - 6270 University Boulevard, Vancouver, BC, Canada V6T 1Z4
\\
\bigskip

@@

\end{flushleft}
% Please keep the abstract below 300 words
%currently 312 words
\section*{Abstract}
Genetic sex determination is remarkably dynamic; many taxa display shifts in the location of sex-determining loci or the evolution of entirely new sex-determining systems. 
Predominant theories for why we observe such transitions generally conclude that novel sex-determining systems are favoured by selection if they equalise the sex ratio or increase linkage with a sexually-antagonistic locus. 
We use population genetic models to extend these theories in two ways: 
(1) We explicitly consider selection on loci very tightly linked to the ancestral sex-determining loci, e.g., within the non-recombining region of the ancestral sex chromosomes. 
Variation at such loci can favour the spread of new sex-determining systems in which the heterogametic sex changes (XY to ZW or ZW to XY) and the new sex-determining region is less closely linked (or unlinked) to the locus under selection, which is not predicted by previous theory. 
(2) We also consider selection upon haploid genotypes either during gametic competition (e.g., pollen/sperm competition involving multiple males) or meiosis (i.e., non-Mendelian segregation), which can cause the zygotic sex ratio to become biased. 
We find that associations with haploid selected loci can drive transitions between sex-determining systems, without requiring sexually-antagonistic selection in diploids. 
Unexpectedly, with haploid selection, transitions between male and female heterogamety can also evolve where linkage with the sex-determining locus is weakened. 
Furthermore, sex ratio biases may increase or decrease with the spread of new sex chromosomes. 
Thus, we find that transitions between sex-determining systems cannot be simply predicted by selection to equalise the sex ratio. 
Overall, our models reveal that transitions between sex-determining systems, particularly transitions where the heterogametic sex changes, can be driven by loci in previously unpredicted genomic locations that experience selection during diploid and/or haploid phases.
These results predict conditions under which sex-determining systems are likely to be labile and draw novel connections with sex ratio evolution.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
%currently 193 words
\section*{Author summary}
Systems of sex determination are strikingly diverse and labile in many clades. 
This poses the question: what drives transitions between sex-determining systems? 
Here, we use models to derive conditions under which new sex-determining systems spread. 
Prevailing views suggest that new sex-determining systems are favoured when they equalize the sex ratio and/or when they are more closely linked to genes that experience differential selection in males and females. 
Our models extend these theories to include selection upon haploid genotypes (meiotic drive or gametic competition), which causes sex ratio biases and occurs differently in males and females. 
Surprisingly, we find that neither force (sex ratio selection nor associations with genes that have sex-specific effects) dominates the spread of new sex-determining systems alone. 
Even more unexpectedly, we find that, to spread, new sex-determining alleles do not necessarily have to arise in closer linkage with genes that are differentially selected in males and females. 
Therefore, our models predict loci in previously unexpected genomic locations and/or experiencing various types of selection (including haploid selection) can now be implicated as drivers of transitions between sex-determining systems.

@@

%MAJOR MECHANISMS
%(1) sex-antagonism
Predominant theories accounting for transitions between GSD systems by selection involve fitness differences between sexes (e.g., sexually antagonistic selection) or sex ratio selection~\cite{Blaser2012, Beukeboom:2014vb,vanDoorn2014re}. 
%\citet{Bull:1977wt} shows that direct selection on new sex determination loci can allow them to spread and switch between male and female heterogamety. 
van Doorn and Kirkpatrick~\cite{vanDoorn:2007eu,vanDoorn:2010hu} and Muralidhar and Veller~\cite{Muralidhar2018} have shown that new sex-determining alleles can be favoured if they arise at loci in close linkage with a locus that experiences sexual antagonism. 
Tighter linkage allows a stronger favourable association to build up between a male-beneficial allele, and a neo-Y allele, for example. 
Such associations can favour cis-GSD transitions~\cite{vanDoorn:2007eu}, trans-GSD transitions~\cite{vanDoorn:2010hu}, and new partially-masculinizing or partially-feminizing alleles~\cite{Muralidhar2018}.
However, any sexually-antagonistic loci that are more closely linked to the ancestral sex-determining locus will develop similar, favourable associations and are expected to hinder the spread of a new sex-determining system.

%MAJOR MECHANISMS
%(2) sex-ratio selection
The sex ratio is directly determined by the sex-determining system, and it has therefore been suggested that sex ratio selection is a dominant force in the evolution of sex determination (e.g., Bull, 1983, p 66-67 ~\cite{Bull:1983vi}; Buekeboom and Perrin, 2014, Chapter 7~\cite{Beukeboom:2014vb}). 
`Fisherian' sex-ratio selection favours a 1:1 zygotic sex ratio when assuming that males and females are equally costly to produce~\cite{Fisher:1930wy,Charnov:1982wg}.
This follows from the fact that, for an autosomal locus, half of the genetic material is inherited from a male and half from a female~\cite{West:2009we}. 
Thus, if the population sex ratio is biased towards one sex, the average per-individual contribution of genetic material to the next generation from the opposite sex is greater. 
Therefore, a mutant that increases investment in the rarer sex will spread via the higher per-individual contributions made by that sex. 
%The default mode of sex ratio selection is `Fisherian' sex ratio selection, which favours equal investment in male and female offspring \citep[i.e., a 1:1 zygotic sex ratio when assuming that males and females are equally costly to produce,][]{Fisher:1930wy,Charnov:1982wg}. 
%Given that the sex-determining system can directly affect the sex ratio, we might expect Fisherian sex ratio selection to influence the spread of new sex-determining systems. 
%In the case of sex-chromosome evolution, Kozielska et al. (2010)~\cite{Kozielska:2010vm} consider systems in which the ancestral sex chromosomes experience meiotic drive (e.g., where driving X or Y chromosomes are inherited disproportionately often), which causes sex ratios to become biased~\cite{Hamilton:1967ts}. 
%They find that masculinizing or feminizing mutations that arise at a new, unlinked locus (i.e., neo-Y or neo-W chromosomes) can then spread, which restore an even sex ratio. 
%However, starting from ESD, Ubeda et al. (2015)~\cite{Ubeda:2015fx} have numerically shown that sex-specific meiotic drive can allow masculinizing and feminizing mutations to spread (i.e., ESD $\rightarrow$ GSD) despite causing a sex-ratio bias.

%HOLES IN CURRENT THEORY
%(1) haploid selection; it exists in many spp
While both opposing selection in the diploid sexes and sex-ratio selection surely influence the evolution of new sex-determining factors, so too might selection in the haploid phase.
Even in animal and plant species that have much larger and more conspicuous diploid phases than haploid phases, many loci experience significant haploid selection through gamete competition and/or meiotic drive~\cite{Mulcahy:1996ha,JOSEPH:2004haa}.
We use the term `meiotic drive' to refer to the biased (non-Mendelian) segregation of genotypes during gamete production (from one parent only) and the term `gametic competition' to refer to selection upon haploid genotypes within a gamete/gametophyte pool (from many parents); the term `haploid selection' encompasses both processes. 
Segregation distortion provides putative evidence of haploid selection and can sometimes be attributed to meiotic drive and/or gametic competition~\cite{Lalanne2004,Fishman2005,Leppala2008,Leppala2013,Didion2015,Didion2016}.
Where it has been characterized, meiotic drive generally occurs either during the production of male or female gametes only~\cite{Ubeda:2005gw,Lindholm:2016cw}.
Gametic competition is also typically sex specific, occurring primarily among male gametes, because there are typically many more pollen/sperm than required for fertilization.
Gametic competition may be particularly common in plants, in which 60-70\% of all genes are expressed in the male gametophyte, and these genes exhibit stronger signatures of selection than randomly-chosen genes~\cite{Borg:2009jpa,Arunkumar:2013iq,Gossmann:2014dua}.
In addition, artificial selection pressures applied to male gametophytes are known to cause a response to selection (e.g.,~\cite{Hormaza:1996cv,Ravikumar:2003uo,Hedhly:2004iv,Clarke:2004ir}). 
A smaller proportion of genes are thought to be expressed and selected during competition in animal sperm, although precise estimates are uncertain~\cite{Zheng:2001fi,JOSEPH:2004haa,Vibranovski:2010et}. 
Nevertheless, recent studies have demonstrated that sperm competition in animals, even within a single ejaculate, can alter haploid allele frequencies and increase offspring fitness~\cite{Immler:2014im,Alavioon2017}.

%HOLES IN CURRENT THEORY
%(2) haploid selection; it can complicate outcomes for SD transitions
There are various ways by which haploid selection could influence transitions between sex-determining systems. 
If we assume that haploid selection at any particular locus predominantly occurs in one sex (e.g., meiotic drive during spermatogenesis), then such loci experience a form of sex-specific selection. 
In this respect, we might expect that haploid selection would affect transitions between sex-determining systems in a similar manner to sex-specific diploid selection (as explored in~\cite{vanDoorn:2007eu,vanDoorn:2010hu}). 
That is, new masculinizing mutations (neo-Y alleles) could be favoured via associations with alleles that are beneficial in the male haploid stage (e.g., \cite{Ubeda:2015fx}). 
Kozielska et al. (2010)~\cite{Kozielska:2010vm} have shown this to be theoretically possible for the special case of perfect linkage between the new sex-determining locus and a locus that experiences meiotic drive in one sex, under a limited number of forms of diploid selection.
Similarly, Ubeda et al. (2015)~\cite{Ubeda:2015fx} numerically show that sex-specific meiotic drive can allow masculinizing and feminizing mutations to spread in an ancestrally environmental-sex-determination system, again under a limited number of forms of diploid selection. 
On the other hand, sex ratios can also become biased by linkage between the sex-determining locus and a locus that harbours genetic variation in haploid fitness, as indeed occurs in the two examples above.
There are also several known empirical cases of sex-ratio bias caused by sex-linked meiotic drive alleles (Burt and Trivers, 2006, Chapter 6~\cite{Burt:2006}) or selection among X- and Y-bearing pollen~\cite{Lloyd:1974tz,Conn:1981uw,Stehlik:2005ul,Stehlik:2006to,Field:2012fd,Field:2013cc}. 
Sex-ratio selection can thus select against the spread of linked sex-determining alleles and favour the invasion of alternative, unlinked sex-determination systems \cite{Kozielska:2010vm,Ubeda:2015fx}.
%It is not immediately clear how the spread of new sex-determining systems would be influenced by the combination of sex ratio biases and associations with haploid selected alleles. 
It is not yet clear when new sex-determining alleles are favoured by associations with haploid selected alleles vs. sex-ratio bias, especially under a wide range of possible diploid selection regimes (e.g., sex-antagonism, overdominance).	
Here we use mathematical models to clarify the conditions under which new sex-determining systems spread when individuals experience extremely general and potentially sex-specific selection at both diploid and haploid stages, which allows fitness differences between the sexes and sex-ratio biases to occur simultaneously.

%MAIN FINDINGS IN GENERAL TERMS
We find that the spread of novel sex-determining systems is influenced by both Fisherian sex-ratio selection and by selection on linked alleles. 
%We find that sex ratio biases caused by haploid selection can exert Fisherian sex ratio selection upon novel sex-determiners but that their spread is also determined by selection on genetically-associated alleles. 
%Indeed, it is only when haploid-selected loci are tightly linked to the ancestral sex-determining region (and so sex ratio biases are initially large) that we see an asymmetry between selection for XY to ZW transitions and ZW to XY transitions (e.g., because haploid selection in males only causes biased zygotic sex ratios in an ancestrally XY system).
Fisherian sex-ratio selection does not always dominate; it is possible for selection on linked alleles to drive turnover between sex-determining systems despite causing increasingly biased sex ratios, and our analytical results elucidate what is required for this to be so.
In addition to considering haploid selection, another novel development in our model is that we consider loci that are in very tight linkage with the ancestral sex-determining locus. 
%Because sex-determining loci are often found within a region of suppressed recombination, there can be a significant number of tightly linked loci. 
We find that loci tightly linked with the sex-determining locus can drive transitions in which the heterogametic sex changes, even when the new sex-determining locus is less closely linked to loci under selection (either including haploid selection or not).

@@

We draw three main points about the generic invasion of neo-Y and neo-W mutations from Table \ref{tab:haplotype_growth}.
First, invasion by a neo-Y (neo-W) does not directly depend on the fitness of female (male) diploids. %(for a given set of equilibrium allele frequencies).
This is because a dominant neo-Y (neo-W) is always found in males (females), and therefore the frequency of the neo-Y (neo-W), $m$, only changes in males (females), Fig \ref{fig:model_outline}C,D.
Second, Fisherian sex-ratio selection will favour the spread of a neo-Y and inhibit the spread of a neo-W if the ancestral zygotic sex ratio is biased towards females (i.e., the first factor of the $\Lambda_{mi}^{(XY)}$ is greater than one for a neo-Y and less than one for a neo-W when $\zeta<1/2$).
However, the spread of a neo-Y (neo-W) also depends on the male (female) fitness of associated alleles (see terms involving equilibrium allele frequencies, $\hat{p}$'s). That is, 
\vspace{0.5cm}

@@

In order to explicitly determine the conditions under which a rare new sex-determining allele spreads, we must calculate the equilibrium frequency of the $A$ allele (i.e., $\hat{p}^\female_X$, $\hat{p}^\male_X$, and $\hat{p}^\male_Y$) and Y-bearing male gametes ($\hat{q}$) in the ancestral population. 
Since only the $\mathbf{A}$ locus experiences selection directly, any deterministic evolution requires that there is a polymorphism at the $\mathbf{A}$ locus. 
Polymorphisms can be maintained by mutation-selection balance or transiently present during the spread of beneficial alleles. 
However, polymorphisms maintained by selection can maintain alleles at intermediate allele frequencies for longer periods. 
Here, we focus on polymorphisms maintained by selection, where the $A$ allele reaches a stable intermediate equilibrium frequency under the ancestral sex-determining system before the new sex-determining allele ($m$) arises. 
Such polymorphisms can be maintained by heterozygote advantage, sexually-antagonistic selection, ploidally-antagonistic selection, or a combination~\cite{Immler:2012tl}.
We can analytically calculate the allele frequency of the $A$ allele using two alternative simplifying assumptions: 
(1) the $\mathbf{A}$ locus is within (or tightly linked to) the non-recombining region around the ancestral sex-determining locus ($r \approx 0$) or (2) selection is weak relative to recombination ($s^\circ$, $t^\circ$, $\alpha_{\Delta}^\circ$ of order $\epsilon<<1$).

>>>>

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Haploid selection, sex-ratio bias, and transitions between sex-determining systems} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Michael Francis Scott\textsuperscript{1\Yinyang*},
Matthew Miles Osmond\textsuperscript{2\Yinyang},
Sarah Perin Otto\textsuperscript{2}
\\
\bigskip
\textbf{1} UCL Genetics Institute, Department of Genetics, Evolution and Environment, University College London, Gower Street, London WC1E 6BT 
\\
\textbf{2} Department of Zoology, University of British Columbia, \#4200 - 6270 University Boulevard, Vancouver, BC, Canada V6T 1Z4
\\
\bigskip

@@

\end{flushleft}
% Please keep the abstract below 300 words
%currently 312 words
\section*{Abstract}
Sex determination is remarkably dynamic; many taxa display shifts in the location of sex-determining loci or the evolution of entirely new sex-determining systems. 
Predominant theories for why we observe such transitions generally conclude that novel sex-determining systems are favoured by selection if they equalise the sex ratio or increase linkage with a sexually-antagonistic locus. 
We use population genetic models to extend these theories in two ways: 
(1) We explicitly consider selection on loci very tightly linked to the ancestral sex-determining loci, e.g., within the non-recombining region of the ancestral sex chromosomes. 
Variation at such loci can favour the spread of new sex-determining systems in which the heterogametic sex changes (XY to ZW or ZW to XY) and the new sex-determining region is less closely linked (or unlinked) to the locus under selection, which is not predicted by previous theory. 
(2) We consider selection upon haploid genotypes either during gametic competition (e.g., pollen/sperm competition involving multiple males) or meiosis (i.e., non-Mendelian segregation), which can cause the zygotic sex ratio to become biased. 
We find that associations with haploid selected loci can drive transitions between sex-determining systems, without requiring sexually-antagonistic selection in diploids. 
Unexpectedly, with haploid selection, transitions between male and female heterogamety can also evolve where linkage with the sex-determining locus is weakened. 
Furthermore, sex-ratio biases may increase or decrease with the spread of new sex chromosomes. 
Thus, we find that transitions between sex-determining systems cannot be simply predicted by selection to equalise the sex ratio. 
Overall, our models reveal that transitions between sex-determining systems, particularly transitions where the heterogametic sex changes, can be driven by loci in previously unpredicted genomic locations that experience selection during diploid and/or haploid phases.
These results predict conditions under which sex-determining systems are likely to be labile and draw connections with sex-ratio evolution.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
%currently 193 words
\section*{Author summary}
Systems of sex determination are strikingly diverse and labile in many clades. 
This poses the question: what drives transitions between sex-determining systems? 
Here, we use models to derive conditions under which new sex-determining systems spread. 
Prevailing views suggest that new sex-determining systems are favoured when they equalize the sex ratio and/or when they are more closely linked to genes that experience differential selection in males and females. 
Our models extend these theories to include selection upon haploid genotypes (meiotic drive or gametic competition), which causes sex-ratio biases and occurs differently in males and females. 
Surprisingly, we find the two forces (sex-ratio selection and associations with genes that have sex-specific effects) will often be equally strong, and thus neither dominates the spread of new sex-determining systems alone. 
Even more unexpectedly, we find that new sex-determining alleles can spread despite being less closely linked to selected loci under a wide range of conditions. 
Therefore, our models predict loci in previously unexpected genomic locations and/or experiencing various types of selection (including haploid selection) can now be implicated as drivers of transitions between sex-determining systems.

@@

%MAJOR MECHANISMS
%(1) sex-antagonism
Predominant theories accounting for transitions between GSD systems by selection involve fitness differences between sexes (e.g., sexually antagonistic selection) or sex-ratio selection~\cite{Blaser2012, Beukeboom:2014vb,vanDoorn2014re}. 
%\citet{Bull:1977wt} shows that direct selection on new sex determination loci can allow them to spread and switch between male and female heterogamety. 
van Doorn and Kirkpatrick~\cite{vanDoorn:2007eu,vanDoorn:2010hu} and Muralidhar and Veller~\cite{Muralidhar2018} have shown that new sex-determining alleles can be favoured if they arise at loci in close linkage with a locus that experiences sexual antagonism. 
Tighter linkage allows a stronger favourable association to build up between a male-beneficial allele, and a neo-Y allele, for example. 
Such associations can favour cis-GSD transitions~\cite{vanDoorn:2007eu}, trans-GSD transitions~\cite{vanDoorn:2010hu}, and new partially-masculinizing or partially-feminizing alleles~\cite{Muralidhar2018}.
However, any sexually-antagonistic loci that are more closely linked to the ancestral sex-determining locus will develop similar, favourable associations and are expected to hinder the spread of a new sex-determining system.

%MAJOR MECHANISMS
%(2) sex-ratio selection
The sex ratio is directly determined by the sex-determining system, and it has therefore been suggested that sex-ratio selection is a dominant force in the evolution of sex determination (e.g., Bull, 1983, p 66-67 ~\cite{Bull:1983vi}; Buekeboom and Perrin, 2014, Chapter 7~\cite{Beukeboom:2014vb}). 
`Fisherian' sex-ratio selection favours a 1:1 zygotic sex ratio when assuming that males and females are equally costly to produce~\cite{Fisher:1930wy,Charnov:1982wg}.
This follows from the fact that, for an autosomal locus, half of the genetic material is inherited from a male and half from a female~\cite{West:2009we}. 
Thus, if the sex ratio is biased, an individual of the rarer sex will, on average, contribute more genetic material to the next generation. 
Selection therefore favours mutants that increase investment in the rarer sex.
%The default mode of sex-ratio selection is `Fisherian' sex-ratio selection, which favours equal investment in male and female offspring \citep[i.e., a 1:1 zygotic sex ratio when assuming that males and females are equally costly to produce,][]{Fisher:1930wy,Charnov:1982wg}. 
%Given that the sex-determining system can directly affect the sex ratio, we might expect Fisherian sex-ratio selection to influence the spread of new sex-determining systems. 
%In the case of sex-chromosome evolution, Kozielska et al. (2010)~\cite{Kozielska:2010vm} consider systems in which the ancestral sex chromosomes experience meiotic drive (e.g., where driving X or Y chromosomes are inherited disproportionately often), which causes sex ratios to become biased~\cite{Hamilton:1967ts}. 
%They find that masculinizing or feminizing mutations that arise at a new, unlinked locus (i.e., neo-Y or neo-W chromosomes) can then spread, which restore an even sex ratio. 
%However, starting from ESD, Ubeda et al. (2015)~\cite{Ubeda:2015fx} have numerically shown that sex-specific meiotic drive can allow masculinizing and feminizing mutations to spread (i.e., ESD $\rightarrow$ GSD) despite causing a sex-ratio bias.

%HOLES IN CURRENT THEORY
%(1) haploid selection; it exists in many spp
While both opposing selection in the diploid sexes and sex-ratio selection surely influence the evolution of new sex-determining factors, so too might selection in the haploid phase.
Even in animal and plant species that have much larger and more conspicuous diploid phases than haploid phases, many loci experience significant haploid selection through gamete competition and/or meiotic drive~\cite{Mulcahy:1996ha,JOSEPH:2004haa}.
We use the term `meiotic drive' to refer to the biased (non-Mendelian) segregation of genotypes during gamete production (from one parent only) and the term `gametic competition' to refer to selection upon haploid genotypes within a gamete/gametophyte pool (from many parents); the term `haploid selection' encompasses both processes. 
Segregation distortion provides putative evidence of haploid selection and can sometimes be attributed to meiotic drive and/or gametic competition~\cite{Lalanne2004,Fishman2005,Leppala2008,Leppala2013,Didion2015,Didion2016}.
Where it has been characterized, individual cases of meiotic drive are generally sex-limited, exclusively affecting male or female gametes~\cite{Ubeda:2005gw,Lindholm:2016cw}.
Gametic competition is also typically sex-limited, occurring primarily among male gametes, because there are typically many more pollen/sperm than required for fertilization.
Gametic competition may be particularly common in plants, in which 60-70\% of all genes are expressed in the male gametophyte, and these genes exhibit stronger signatures of selection than randomly-chosen genes~\cite{Borg:2009jpa,Arunkumar:2013iq,Gossmann:2014dua}.
In addition, artificial selection pressures applied to male gametophytes are known to cause a response to selection (e.g.,~\cite{Hormaza:1996cv,Ravikumar:2003uo,Hedhly:2004iv,Clarke:2004ir}). 
A smaller proportion of genes are thought to be expressed and selected during competition in animal sperm, although precise estimates are uncertain~\cite{Zheng:2001fi,JOSEPH:2004haa,Vibranovski:2010et}. 
Nevertheless, recent studies have demonstrated that sperm competition in animals, even within a single ejaculate, can alter haploid allele frequencies and increase offspring fitness~\cite{Immler:2014im,Alavioon2017}.

%HOLES IN CURRENT THEORY
%(2) haploid selection; it can complicate outcomes for SD transitions
There are various ways by which haploid selection could influence transitions between sex-determining systems. 
If we assume that haploid selection at any particular locus predominantly occurs in one sex (e.g., meiotic drive during spermatogenesis), then such loci experience a form of sex-specific selection. 
In this respect, we might expect that haploid selection would affect transitions between sex-determining systems in a similar manner to sex-specific diploid selection (as explored in~\cite{vanDoorn:2007eu,vanDoorn:2010hu}). 
That is, new masculinizing mutations (neo-Y alleles) could be favoured via associations with alleles that are beneficial in the male haploid stage (e.g., \cite{Ubeda:2015fx}). 
Kozielska et al. (2010)~\cite{Kozielska:2010vm} have shown this to be theoretically possible for the special case of perfect linkage between the new sex-determining locus and a locus that experiences meiotic drive in one sex, under a limited number of forms of diploid selection.
Similarly, Ubeda et al. (2015)~\cite{Ubeda:2015fx} numerically show that sex-specific meiotic drive can allow masculinizing and feminizing mutations to spread in an ancestrally environmental-sex-determination system, again under a limited number of forms of diploid selection. 
On the other hand, sex ratios can also become biased by linkage between the sex-determining locus and a locus that harbours genetic variation in haploid fitness, as indeed occurs in the two examples above.
There are also several known empirical cases of sex-ratio bias caused by sex-linked meiotic drive alleles (Burt and Trivers, 2006, Chapter 6~\cite{Burt:2006}) or selection among X- and Y-bearing pollen~\cite{Lloyd:1974tz,Conn:1981uw,Stehlik:2005ul,Stehlik:2006to,Field:2012fd,Field:2013cc}. 
Sex-ratio selection can thus select against the spread of linked sex-determining alleles and favour the invasion of alternative, unlinked sex-determination systems \cite{Kozielska:2010vm,Ubeda:2015fx}.
%It is not immediately clear how the spread of new sex-determining systems would be influenced by the combination of sex-ratio biases and associations with haploid selected alleles. 
It is not yet clear when new sex-determining alleles are favoured by associations with haploid selected alleles vs. sex-ratio bias, especially under a wide range of possible diploid selection regimes (e.g., sex-antagonism, overdominance).

` 
We consider both selected loci that are loosely linked to the sex-determining region (e.g., on an autosome) and selected loci that are in very tight linkage with the ancestral sex-determining locus (e.g., within the non-recombining sex-determining region).
Our main findings are three:
(1) transitions between sex-determining systems can decrease or increase sex-ratio bias, and with sufficiently loose linkage between the selected locus and the sex-determining locus, the selective force exerted by sex-ratio selection and favourable associations with haploid-selected alleles are exactly equivalent;
(2) ancestrally-tight sex-linkage and haploid selection allow trans-GSD transitions that weaken sex-linkage; and
(3) haploid selection removes the need for sex-antagonism, increasing the range of scenarios that select for transitions between sex-determining systems.
%MAIN FINDINGS IN GENERAL TERMS
%We find that the spread of novel sex-determining systems is influenced by both Fisherian sex-ratio selection and by selection on linked alleles. 
%We find that sex-ratio biases caused by haploid selection can exert Fisherian sex-ratio selection upon novel sex-determiners but that their spread is also determined by selection on genetically-associated alleles. 
%Indeed, it is only when haploid-selected loci are tightly linked to the ancestral sex-determining region (and so sex-ratio biases are initially large) that we see an asymmetry between selection for XY to ZW transitions and ZW to XY transitions (e.g., because haploid selection in males only causes biased zygotic sex ratios in an ancestrally XY system).
%Fisherian sex-ratio selection does not always dominate; it is possible for selection on linked alleles to drive turnover between sex-determining systems despite causing increasingly biased sex ratios, and our analytical results elucidate what is required for this to be so.
%Because sex-determining loci are often found within a region of suppressed recombination, there can be a significant number of tightly linked loci. 
%We find that loci tightly linked with the sex-determining locus can drive transitions in which the heterogametic sex changes, even when the new sex-determining locus is less closely linked to loci under selection (either including haploid selection or not).

@@

We draw three main points about the generic invasion of neo-Y and neo-W mutations from Table \ref{tab:haplotype_growth}.
First, invasion by a neo-Y (neo-W) does not directly depend on the fitness of female (male) diploids. %(for a given set of equilibrium allele frequencies).
This is because a dominant neo-Y (neo-W) is always found in males (females), and therefore the frequency of the neo-Y (neo-W), $m$, only changes in males (females), Fig \ref{fig:model_outline}C,D.
Second, Fisherian sex-ratio selection will favour the spread of a neo-W and inhibit the spread of a neo-Y if the ancestral zygotic sex ratio is biased towards males (i.e., the first factor of the $\Lambda_{mi}^{(XY)}$ is greater than one for a neo-W and less than one for a neo-Y when $\zeta>1/2$), as might occur when the ancestral sex-determining locus is linked to a locus experiencing meiotic drive in males, for example.
However, the spread of a neo-Y (neo-W) also depends on the male (female) fitness of associated alleles (see terms involving equilibrium allele frequencies, $\hat{p}$'s). That is, 
\vspace{0.5cm}

@@

In order to explicitly determine the conditions under which a rare new sex-determining allele spreads, we must calculate the equilibrium frequency of the $A$ allele (i.e., $\hat{p}^\female_X$, $\hat{p}^\male_X$, and $\hat{p}^\male_Y$) and Y-bearing male gametes ($\hat{q}$) in the ancestral population. 
Since only the $\mathbf{A}$ locus experiences selection directly, any deterministic evolution requires that there is a polymorphism at the $\mathbf{A}$ locus. 
Polymorphisms can be maintained by mutation-selection balance or transiently present during the spread of beneficial alleles. 
However, polymorphisms maintained by selection can maintain alleles at intermediate allele frequencies for longer periods. 
Here, we focus on polymorphisms maintained by selection, where the $A$ allele reaches a stable intermediate equilibrium frequency under the ancestral sex-determining system before the new sex-determining allele ($m$) arises. 
Such polymorphisms can be maintained by heterozygote advantage, sexually-antagonistic selection, ploidally-antagonistic selection, or a combination~\cite{Immler:2012tl}.
We can analytically calculate the allele frequency of the $A$ allele using two alternative simplifying assumptions: 
(1) the $\mathbf{A}$ locus is within (or tightly linked to) the non-recombining region around the ancestral sex-determining locus ($r \approx 0$) or (2) selection is weak relative to recombination ($s^\circ$, $t^\circ$, $\alpha_{\Delta}^\circ$ $<<$ $r$).
